dyngen: benchmarking with \textit{in silico} single cells

### 2019-10-10 content.tex
update manuscript

We introduce \texttt{dyngen}, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:dyngen}A). A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} where a cell consists of a set of molecules. Throughout a simulation, reactions (e.g. transcription, splicing) modify the abundance of these molecules, and the likelihood of a reaction occurring is again dependent on molecule abundance.

By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:dyngen}B). We demonstrate \texttt{dyngen}'s flexibility by simulating three different types of biological experiments, and using these simulations to develop benchmarking methodologies for recently developed computational tools.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/showcase_4} 
	\caption{
		\textbf{Showcase of dyngen functionality.} 
		\textbf{A:} 
		\textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
	}
	\label{fig:dyngen}
\end{figure}

@@

One of dyngen's main advantages is that depending on how the GRN is designed, different cellular developmental processes can be simulated. Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies. Multiple simulations with slightly different GRNs can emulate rewiring events in disease or batch effects.

\mycomment{could use figure; one with GRNs of different topologies, another with rewiring events}

Another advantage is that dyngen returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F). These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches. For example, a network inference method could use mRNA abundance and cellular states as inputs, and its output could be benchmarked against the gold standard GRN.

The final main advantage is that \texttt{dyngen} can be used to simulate different experimental conditions. By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation). 
It is possible to implement other experimental protocols, such as sampling the same cell at regular intervals or performing perturbation experiments.

\mycomment{add result figures pertaining use cases}

\section{Discussion}
As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications. This may include methods that integrate clustering, network inference and trajectory inference. In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.

Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other. Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.  Sampling a cell at a particular time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

dyngen ultimately also allows anticipating technological developments in single-cell multi-omics. In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
Similarly, it could also be used to compare which experimental technique will likely produce the most accurate result. For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knockdown experiments a necessity in order to infer high-quality regulatory networks?

Currently, dyngen focuses on simulating cells as standalone entities.
Future developments include extending the framework to simulate multiple cells in a virtual environment. Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.

%\mycomment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\hugefigure]{fig/explain_methods} 
	\caption{\textbf{The workflow of dyngen is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network. \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the mocules at each time step are extracted. Each time step is mapped to a point in the backbone. \textbf{F:} Multiple cells are sampled from each simulation. Molecules are sampled from each cell.}
	\label{fig:explain_methods}
\end{figure}

@@

In dyngen, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules. A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen_ssa}). At the end of the dyngen process, cells will not be sampled from the burn phase simulations.

@@

\mycomment{mention strength, cooperativity, effect and independence factor.}

@@

\begin{table}[ht]
	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
	\centering
	\begin{tabular}{|lcc|}
		\hline
		Reaction & Effect & Propensity \\ \hline \hline
		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \reg{G,H})}$ \\
		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
		mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
		Translation & $\x G \rightarrow \w G + \y G$ & $\ypr G \times \x G$ \\
		Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
	\end{tabular}
\end{table}
%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
% TODO: add step-by-step  derivation of transcription formula
%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

@@

\begin{table}[ht]
	\caption{\textbf{Parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
	\centering
	\begin{tabular}{|lrl|}
		\hline
		Parameter & Symbol & Definition \\ \hline \hline
		Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
		Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
		mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
		Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
		Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
		Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\ 
		Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
    Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
		TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\ 
		Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
		Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
		\multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
	\end{tabular}
\end{table}

\subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state. It is assumed the root state has no modules active and has no expression of any molecules. To get to next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition. To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.

@@

\texttt{dyngen} uses Gillespie's Stochastic Simulation Algorithm (SSA) to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

@@

\subsubsection{Map SSA simulations to backbone}
The cellular state of each timepoint in the SSA simulation is mapped to the state network of the backbone by calculating the 1NN between a state vector in the simulation and the average expression levels along transitions.

@@

\subsection{Example runs of predefined backbones}

\subsubsection{Linear}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{fig/example/bb_linear/plot} 
	\caption{}
	\label{fig:example_linear}
\end{figure}

\subsubsection{Bifurcating}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating/plot} 
	\caption{}
	\label{fig:example_bifurcating}
\end{figure}

\subsubsection{Cycle}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{fig/example/bb_cycle/plot} 
	\caption{}
	\label{fig:example_cycle}
\end{figure}

%\subsubsection{Branching}
%
%\begin{figure}[H]
%	\centering
%	\includegraphics[width=\linewidth]{fig/example/bb_binary_tree/plot} 
%	\caption{}
%	\label{fig:example_branching}
%\end{figure}

\subsubsection{Converging}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{fig/example/bb_converging/plot} 
	\caption{}
	\label{fig:example_converging}
\end{figure}

\subsubsection{Bifurcating converging}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_converging/plot} 
	\caption{}
	\label{fig:example_bifurcating_converging}
\end{figure}

%\subsubsection{Bifurcating cycle}
%
%\begin{figure}[H]
%	\centering
%	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_cycle/plot} 
%	\caption{}
%	\label{fig:example_bifurcating_cycle}
%\end{figure}

\subsubsection{Bifurcating loop}

\begin{figure}[H]
	\centering
	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_loop/plot} 
	\caption{}
	\label{fig:example_bifurcating_loop}
\end{figure}

%\subsubsection{Disconnected}
%
%\begin{figure}[H]
%	\centering
%	\includegraphics[width=\linewidth]{fig/example/bb_disconnected/plot} 
%	\caption{}
%	\label{fig:example_disconnected}
%\end{figure}

\subsection{Example use cases}
\subsubsection{Trajectory alignment}
From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other. 
\subsubsection{Differential network inference}
From discussion: Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
\subsubsection{RNA velocity}
From discussion: Sampling a cell at a certain time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

\subsubsection{Perturbation experiment}

>>>>

We introduce \texttt{dyngen}, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:showcase}A). A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} where a cell consists of a set of molecules. Throughout a simulation, reactions (e.g. transcription, splicing) modify the abundance of these molecules, and the likelihood of a reaction occurring is again dependent on molecule abundance.

By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:showcase}B). We demonstrate \texttt{dyngen}'s flexibility by simulating three different types of biological experiments, and using these simulations to develop benchmarking methodologies for recently developed computational tools.

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/showcase_4} 
	\caption{
		\textbf{Showcase of \texttt{dyngen} functionality.} 
		\textbf{A:} 
		\textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
	}
	\label{fig:showcase}
\end{figure}

@@

\texttt{dyngen} returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F). These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches. For example, a network inference method could use mRNA abundance and cellular states as inputs, and its output could be benchmarked against the gold standard GRN.

Depending on how the GRN is designed, different cellular developmental processes can be simulated.
\texttt{dyngen} includes generators of GRNs which result in many different developmental topologies (Figure~\ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.

\begin{figure}[ht]
	\centering
	\includegraphics[width=\LARgefigure]{fig/example_runs_2} 
	\caption{
		\textbf{Multiple executions of \texttt{dyngen} with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled. 
		\textbf{A:}~Linear. \textbf{B:}~Bifurcating. \textbf{C:}~Converging.
		\textbf{D:}~Cyclic. \textbf{E:}~Bifurcating loop. \textbf{F:}~Bifurcating converging.
		\textbf{G:}~Consecutive branching. \textbf{H:}~Binary tree. \textbf{I:}~Disconnected.
	}
	\label{fig:example_runs}
\end{figure}

Custom-defined GRNs offer more fine-grained control over the simulation.
Aside from simulating topologies currently not supported by \texttt{dyngen}, this has several other important use-cases. Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure~\ref{fig:usecases}). 
Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.

\begin{figure}[ht]
	\centering
	\includegraphics[width=\LARgefigure]{fig/usecases} 
	\caption{
		\textbf{Examples of simulations with perturbed GRNs.}  
		\textbf{A:} The cells in the top half were simulated with the same GRN as the cells on the bottom half, except that all parameters (e.g. strength of the interaction) were randomised.
		\textbf{B:} Only 10 interactions in the GRN were randomised. In this example, the effect of the GRN perturbation is more subtle.
	}
	\label{fig:usecases}
\end{figure}

\texttt{dyngen} can be used to simulate different experimental conditions. By default, \texttt{dyngen} supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation). 
However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point. This would allow evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

%\subsection{Example use cases}
%\subsubsection{Trajectory alignment}
%\subsubsection{Differential network inference}
%\subsubsection{RNA velocity}
%\subsubsection{Perturbation experiment}

\section{Discussion}
As is, \texttt{dyngen}'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications. This may include methods that integrate clustering, network inference and trajectory inference. In this respect, \texttt{dyngen} may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.

\texttt{dyngen} ultimately allows anticipating technological developments in single-cell multi-omics. In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing the qualitative and robust results in downstream analysis.

Currently, \texttt{dyngen} focuses on simulating cells as standalone entities that are well mixed.
Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration\cite{smith_spatialstochasticintracellular_2019}.

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\hugefigure]{fig/explain_methods} 
	\caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network. \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states. The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge. \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.  The counts of each of the mocules at each time step are extracted. Each time step is mapped to a point in the backbone. \textbf{F:} Multiple cells are sampled from each simulation. Molecules are sampled from each cell.}
	\label{fig:explain_methods}
\end{figure}

@@

In \texttt{dyngen}, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules. A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen_ssa}). At the end of the \texttt{dyngen} process, cells will not be sampled from the burn phase simulations. Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).

@@

@@

%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
% TODO: add step-by-step  derivation of transcription formula
%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!

% YSA: Graag wat meer uitleg in de tekst over al deze formules, definitie van notaties en zo.

\begin{table}[ht]
	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$. Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
	\centering
	\begin{tabular}{|lcc|}
		\hline
		Reaction & Effect & Propensity \\ \hline \hline
		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \reg{G,H})}$ \\
		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
		mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
		Translation & $\x G \rightarrow \w G + \y G$ & $\ypr G \times \x G$ \\
		Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
	\end{tabular}
\end{table}

@@

\begin{table}[ht]
	\caption{\textbf{Default parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
	\centering
	\begin{tabular}{|lrl|}
		\hline
		Parameter & Symbol & Definition \\ \hline \hline
		Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
		Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
		mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
		Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
		Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
		Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\ 
		Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
    Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
		TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\ 
		Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
		Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
		\multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
	\end{tabular}
\end{table}

\subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state. It is assumed the root state has no modules active and has no expression of any molecules. To get to the next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition. To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.

% mention that this is basically an ode?

@@

\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA)\cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.

@@

\subsubsection{Map SSA simulations to backbone}
We compute the Pearson correlation between the state vectors in the simulation and the average expression levels along a transition in the backbone. Each timepoint in the SSA simulation is mapped to the point in the backbone that has the highest correlation value.

@@
